### 2013-04-30 doc/dissertation/document/chapters/mc_background/mc_background.tex
adding expected value estimator variance

\subsection{Collision Estimator Variance}
\label{subsec:forward_variance}
We can compute the variance of the collision estimator in the same way
as the direct estimator for each component in the solution where now:
\begin{equation}
  \sigma_j^2 = \sum_{\nu} P_{\nu} \sum_{m=0}^k W_{m}^2 \delta_{i_m,j} + 2
  \sum_{\nu} P_{\nu} \sum_{\substack{ l=0 \\ m<l}}^k W_m W_l
  \delta_{i_m,j} \delta_{i_l,j} - x_i^2\:.
  \label{eq:collision_variance_1}
\end{equation}
In this case, in the second summation $m \neq l$ for all states due to
the form of the summation and therefore the delta functions,
$\delta_{i_m,j}$ or $\delta_{i_l,j}$, will only be nonzero for random
walks that are in the current solution state (when $i_m = i_l = j$)
and other states visited do not contribute to the variance of the
current state (when $i_m \neq i_l$). This is important as it shows a
decoupling of the variance among the solution vector states, meaning
that the variance in solution state $i$ does not depend on the
variance in solution state $j$.

Expanding the transition probabilities in the first sum again yields a
Neumman summation in the same form as that explored for the forward
estimator, this time with the terms in reverse order and the
introduction of the Kronecker delta:
\begin{equation}
  \sigma_j^2 = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
  \sum_{i_k}^{N} p_{i_k,i_{k-1}}\ldots p_{i_2,i_1} p_{i_1,i_0}
  w^2_{i_k,i_{k-1}}\ldots w^2_{i_2,i_1} w^2_{i_1,i_0}
  b^2_{i_0}\delta_{i_k,j} - x_i^2\:.
  \label{eq:collision_variance_2}
\end{equation}
If we again define $\ve{G} = \ve{P} \circ \ve{W} \circ \ve{W}$, we
then have:
\begin{equation}
  \sigma^2_j = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
  \sum_{i_k}^{N}g_{i_{k},i_{k-1}} \ldots
  g_{i_2,i_1}g_{i_1,i_0}b_{i_0}\delta_{i_k,j} - x_i^2\:,
\label{eq:collision_variance_3}
\end{equation}
where now $\ve{G}$ in this case is the transpose of that used in
Eq~(\ref{eq:direct_variance_6}). The Kronecker delta again implies
that the variance contribution from each random walk will only be in
its current state and for many random walks, many starting weights of
$b_{i_0}$ will contribute to the variance in the same way as they
contribute to the solution tally. As the transpose is formed and the
Neumann series of $\ve{G}$ is in reverse order in
Eq~(\ref{eq:collision_variance_3}) relative to its formulation in the
forward estimator, then Eq~(\ref{eq:collision_variance_3}) and
Eq~(\ref{eq:direct_variance_6}) are equivalent and therefore the
collision estimator is bound by the same restrictions on
$\rho(\ve{G})$ and $\rho(\ve{H})$ to ensure that the variance is
finite.

@@

\subsubsection{Expected Value Estimator Variance}
\label{subsubsec:expected_value_estimator_variance}

>>>>

\subsection{Collision Estimator Variance}
\label{subsec:forward_variance}
We can compute the variance of the collision estimator in the same way
as the direct estimator for each component in the solution where now:
\begin{equation}
  \sigma_j^2 = \sum_{\nu} P_{\nu} \sum_{m=0}^k W_{m}^2 \delta_{i_m,j} + 2
  \sum_{\nu} P_{\nu} \sum_{\substack{ l=0 \\ m<l}}^k W_m W_l
  \delta_{i_m,j} \delta_{i_l,j} - x_j^2\:.
  \label{eq:collision_variance_1}
\end{equation}
In this case, in the second summation $m \neq l$ for all states due to
the form of the summation and therefore the delta functions,
$\delta_{i_m,j}$ or $\delta_{i_l,j}$, will only be nonzero for random
walks that are in the current solution state (when $i_m = i_l = j$)
and other states visited do not contribute to the variance of the
current state (when $i_m \neq i_l$). This is important as it shows a
decoupling of the variance among the solution vector states, meaning
that the variance in solution state $i$ does not depend on the
variance in solution state $j$.

Expanding the transition probabilities in the first sum again yields a
Neumman series in the same form as that explored for the forward
estimator, this time with the terms in reverse order and the
introduction of the Kronecker delta:
\begin{equation}
  \sigma_j^2 = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
  \sum_{i_k}^{N} p_{i_k,i_{k-1}}\ldots p_{i_2,i_1} p_{i_1,i_0}
  w^2_{i_k,i_{k-1}}\ldots w^2_{i_2,i_1} w^2_{i_1,i_0}
  b^2_{i_0}\delta_{i_k,j} - x_j^2\:.
  \label{eq:collision_variance_2}
\end{equation}
If we again define $\ve{G} = \ve{P} \circ \ve{W} \circ \ve{W}$, we
then have:
\begin{equation}
  \sigma^2_j = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
  \sum_{i_k}^{N}g_{i_{k},i_{k-1}} \ldots
  g_{i_2,i_1}g_{i_1,i_0}b_{i_0}\delta_{i_k,j} - x_j^2\:,
\label{eq:collision_variance_3}
\end{equation}
where now $\ve{G}$ in this case is the transpose of that used in
Eq~(\ref{eq:direct_variance_6}). The Kronecker delta again implies
that the variance contribution from each random walk will only be in
its current state and for many random walks, many starting weights of
$b_{i_0}$ will contribute to the variance in the same way as they
contribute to the solution tally. As the transpose is formed and the
Neumann series of $\ve{G}$ is in reverse order in
Eq~(\ref{eq:collision_variance_3}) relative to its formulation in the
forward estimator, then Eq~(\ref{eq:collision_variance_3}) and
Eq~(\ref{eq:direct_variance_6}) are equivalent and therefore the
collision estimator is bound by the same restrictions on
$\rho(\ve{G})$ and $\rho(\ve{H})$ to ensure that the variance is
finite.

@@

\subsubsection{Expected Value Estimator Variance}
\label{subsubsec:expected_value_estimator_variance}
Following the collision estimator variance, the expected value
estimator variance is given as:
\begin{multline}
  \sigma_j^2 = \sum_{\nu} P_{\nu} b_j^2 + 2 \sum_{\nu} P_{\nu} b_j
  \sum_{m=0}^k W_{m} h_{j,i_m} +\\ \sum_{\nu} P_{\nu} \sum_{m=0}^k
  W_{m}^2 h_{j,i_m}^2 + 2 \sum_{\nu} P_{\nu} \sum_{\substack{ l=0
      \\ m<l}}^k W_m W_l h_{j,i_m} h_{j,i_l} - x_j^2\:.
  \label{eq:expected_value_variance_1}
\end{multline}
As the delta functions are no longer present, the final summation in
Eq~(\ref{eq:expected_value_variance_1}) contains contributions for all
valid state combinations in the same and therefore the variance of
each state in the solution tally is coupled to a group of other states
as defined by the sparsity pattern of the iteration. This coupling of
variances is expected due to the deterministic averaging used to
generate the expected value estimator.
